---
title: "Distribution of grants approved by the City of Toronto"
author: "James Bao"
date: 2020-09-24
bibliography: 'references.bib'
categories: ["R"]
tags: ["R Markdown", "plot", "Toronto"]
---
# Abstract

There are millions of dollars provided by the city of Toronto. By analyzing
  the how grants were distributed to different organizations throughout the city of
  Toronto, patterns of distribution patterns emerged. The city of Toronto consistently
  prioritizes programs such as Toronto Foundation of Student Success through the Student
  Nutrition program, as well as other organizations. Increase in total funding allocated
  every year did not create a more diverse range of city funding.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)

#install.packages("tidyverse")
#install.packages('blogdown') 

library(opendatatoronto)
library(dplyr)
library(tidyverse)
library(ggplot2)
require(scales)

```

```{r, include=FALSE}
#import all resources from opendatatoronto
list_package_resources("6f20e59a-1dd9-4a1d-ae9b-ba56a126c7f5")


#community grants 2010-2019

resource_2019 <- get_resource("6061bade-5dc1-48ac-a350-8d06a729f16f")
resource_2018 <- get_resource("a97c8c80-ec87-4f93-b17a-7577d6f0454f")
resource_2017 <- get_resource("0a3e08b4-1eff-4b25-b141-c66eba8995db")
resource_2016 <- get_resource("780e6ddb-4456-4ba8-8752-b61004174624")
resource_2015 <- get_resource("824273e2-31bb-4304-9f1e-9aac3a916dee")
resource_2014 <- get_resource("780e6ddb-4456-4ba8-8752-b61004174624")
resource_2013 <- get_resource("4e12313e-fbd4-4dbe-a95a-8d71f05cd63d")
resource_2012 <- get_resource("edf07ceb-7fc2-4189-b5a5-0b7ae5aa1026")
resource_2011 <- get_resource("4445cd61-20fa-479b-9880-aafda2a80eb5")
resource_2010 <- get_resource("aecba158-1063-475c-8bf6-1a0109ead0b6")
#str(resource_2019)
#str(resource_2018)
#str(resource_2017)
#str(resource_2016)
#str(resource_2015)
#str(resource_2014)
#str(resource_2013)
#str(resource_2012)
#str(resource_2011)
#str(resource_2010)
```

# Introduction

City funding is always a major point of contention every year as the city struggles to fund various programs and organizations all vying for more funding. The City of Toronto approves more and more community grants in order to fund to tackle issues such as community involvement, poverty, diversity and inclusion. I found that the city allocates a majority of annual community grants towards a select few organizations. 

By breaking down the funding distribution to gain an insight on which group benefits the most from the increased grants. It was found that as more funds are made available, a select few organizations such as the Toronto Foundation for Student Success gain an overwhelming portion of the new funds. Close to 43% of city funding in 2019 was given to just 20 of over 800 qualifying applicants. 
This blog will go through a brief overview of the data followed by analysis of individual graphs as well as an overall analysis. However this blog lacks depth due to inconsistent data set formats, which will be improved in future editions. 
  
  
  
# Data

```{r, include=FALSE}
#parse and clean up data

#get relevant data for 2018
funding_2018 <- resource_2018[['TGCC compilation']]
funding_2018 <- data.frame(funding_2018)



#get relevant data 2019
funding_2019 <- resource_2019[['Grants Funded']]
funding_2019 <- head(funding_2019,1000)
funding_2019 <- data.frame(funding_2019)


#get relevant data for 2014
funding_2014 <- resource_2014
funding_2014 <- data.frame(funding_2014)

#remove unwanted rows 
funding_2014 <- funding_2014[!grepl("TOTAL", funding_2014$Organization),]
funding_2014 <- funding_2014[!grepl("Number of Grants", funding_2014$Organization),] 
#rename column for consistency
funding_2014 <- funding_2014 %>% rename(Total.Funding.Amount = Total.Allocations)

#get relevant data 2010
funding_2010 <- resource_2010
funding_2010 <- data.frame(funding_2010)

#remove unwanted rows
funding_2010 <- funding_2010[!grepl("TOTAL", funding_2010$Organization),]
#rename column for consistency
funding_2010 <- funding_2010 %>% rename(Total.Funding.Amount = Total.Allocations)

#get relevant data for 2017
funding_2017 <- data.frame(resource_2017)

#get relevant data for 2016
funding_2016 <- data.frame(resource_2016)
funding_2016 <- funding_2016[!grepl("TOTAL", funding_2016$Organization),]
funding_2016 <- funding_2016 %>% rename(Total.Funding.Amount = Total.Allocations)

#get relevant data for 2015
funding_2015 <- data.frame(resource_2015)
funding_2015 <- funding_2015[!grepl("TOTAL", funding_2015$Organization),]

funding_2015 <- funding_2015 %>% rename(Total.Funding.Amount = Total.Allocations)

#get relevant data for 2013
funding_2013 <- data.frame(resource_2013)
funding_2013 <- funding_2013[!grepl("TOTAL", funding_2013$Organization),]

funding_2013 <- funding_2013 %>% rename(Total.Funding.Amount = Total.Allocations)

#get relevant data for 2012
funding_2012 <- data.frame(resource_2012)
funding_2012 <- funding_2012[!grepl("TOTAL", funding_2012$Organization),]
funding_2012 <- funding_2012 %>% rename(Total.Funding.Amount = Total.Allocations)

#get relevant data for 2011
funding_2011 <- data.frame(resource_2011)
funding_2011 <- funding_2011[!grepl("TOTAL", funding_2011$Organization),]
funding_2011 <- funding_2011 %>% rename(Total.Funding.Amount = Total.Allocations)

```

```{r, include=FALSE}
#function to find the top n organizations who received the most funding
top_n_receivers <- function(funding, n=20){ 
  # extract relevant columns
  simplified<-funding%>%select(Organization, Total.Funding.Amount) 
  #sum up rows of same organization
  simplified<-with(simplified, aggregate(list(Total.Funding.Amount=Total.Funding.Amount), list(Organization=Organization),sum))
  #find top n
  simplified<-top_n(simplified, n ,wt=Total.Funding.Amount)
  return (simplified)

}

#Function to graph the top receivers
graph_top_receivers <- function(data, title){
  ggplot(data,aes(y=reorder(Organization,Total.Funding.Amount),x=Total.Funding.Amount))+geom_bar(stat = "identity") + ggtitle(title) + scale_x_continuous(labels = comma) +xlab("Grants Recieved ($)") + ylab("Organization")
}

#function to calculated total funding granted in a given data frame
total_funded<-function(funding){
  filtered<-filter(funding, Total.Funding.Amount!='NA') %>% select(Total.Funding.Amount)
  return (sum(filtered$Total.Funding.Amount))
}

#function to count 
total_organization_supported <- function(funding){
  return(length(unique(funding$Organization)))
}
```

```{r, include=FALSE}

#consolidating data frames for graphs
top_20_receivers_2019<-top_n_receivers(funding_2019)
top_20_receivers_2018<-top_n_receivers(funding_2018)
top_20_receivers_2017<-top_n_receivers(funding_2017)
top_20_receivers_2016<-top_n_receivers(funding_2016)
top_20_receivers_2015<-top_n_receivers(funding_2015)
top_20_receivers_2014<-top_n_receivers(funding_2014)
top_20_receivers_2013<-top_n_receivers(funding_2013)
top_20_receivers_2012<-top_n_receivers(funding_2012)
top_20_receivers_2011<-top_n_receivers(funding_2011)
top_20_receivers_2010<-top_n_receivers(funding_2010)

total_2010_funding <- total_funded(funding_2010)
total_2011_funding <- total_funded(funding_2011)
total_2012_funding <- total_funded(funding_2012)
total_2013_funding <- total_funded(funding_2013)
total_2014_funding <- total_funded(funding_2014)
total_2015_funding <- total_funded(funding_2015)
total_2016_funding <- total_funded(funding_2016)
total_2017_funding <- total_funded(funding_2017)
total_2018_funding <- total_funded(funding_2018)
total_2019_funding <- total_funded(funding_2019)


total_top_20_2010 <- total_funded(top_20_receivers_2010)
total_top_20_2011 <- total_funded(top_20_receivers_2011)
total_top_20_2012 <- total_funded(top_20_receivers_2012)
total_top_20_2013 <- total_funded(top_20_receivers_2013)
total_top_20_2014 <- total_funded(top_20_receivers_2014)
total_top_20_2015 <- total_funded(top_20_receivers_2015)
total_top_20_2016 <- total_funded(top_20_receivers_2016)
total_top_20_2017 <- total_funded(top_20_receivers_2017)
total_top_20_2018 <- total_funded(top_20_receivers_2018)
total_top_20_2019 <- total_funded(top_20_receivers_2019)

total_organization_2010 <- total_organization_supported(funding_2010)
total_organization_2011 <- total_organization_supported(funding_2011)
total_organization_2012 <- total_organization_supported(funding_2012)
total_organization_2013 <- total_organization_supported(funding_2013)
total_organization_2014 <- total_organization_supported(funding_2014)
total_organization_2015 <- total_organization_supported(funding_2015)
total_organization_2016 <- total_organization_supported(funding_2016)
total_organization_2017 <- total_organization_supported(funding_2017)
total_organization_2018 <- total_organization_supported(funding_2018)
total_organization_2019 <- total_organization_supported(funding_2019)

total_funding_by_year <- data.frame(year = c(2010:2019),
                          funding = c(total_2010_funding, total_2011_funding,
                            total_2012_funding, total_2013_funding,
                            total_2014_funding, total_2015_funding,
                            total_2016_funding, total_2017_funding,
                            total_2018_funding, total_2019_funding),
                          funding_top_20 = c(total_top_20_2010, total_top_20_2011,
                            total_top_20_2012, total_top_20_2013,
                            total_top_20_2014, total_top_20_2015,
                            total_top_20_2016, total_top_20_2017,
                            total_top_20_2018, total_top_20_2019)
                          )
total_funding_by_year<- transform(total_funding_by_year, 
                                  percentage = funding_top_20/funding, 
                                  remaining_funding = funding-funding_top_20
                                  )
       

total_organization_by_year <- data.frame(year = c(2010:2019),
                               organizations = c(
                                 total_organization_2010,total_organization_2011,
                                 total_organization_2012,total_organization_2013,
                                 total_organization_2014,total_organization_2015,
                                 total_organization_2016,total_organization_2017,
                                 total_organization_2018,total_organization_2019
                                 
                               ))

year <- total_funding_by_year$year
total <- total_funding_by_year$funding
top_20 <- total_funding_by_year$funding_top_20
remaining<-total_funding_by_year$remaining_funding
total_funding_by_year_stacked<-rbind(
  data.frame(year, funding=top_20/total*100, type="Total Funding for Top 20"),
  data.frame(year, funding=remaining/total*100,type="Remaining Funding")
  )

total_funding_by_year <- rbind(    
  data.frame(year, funding=top_20, type="Total Funding for Top 20"),
  data.frame(year, funding=total-top_20,type="Total Funding")
  )
```

Every year, the city of Toronto allocates tens of millions of dollars to support the various organizations throughout the city in order to enhance the living experience of its citizens through the . As a member of one of these organizations who vies city grants every year, I know the competitive and lengthy process of obtaining even a few hundred of the millions of dollars given out every year. Luckily, the city of Toronto provided detailed and informative data sets regarding where the money is sent out to every year. The data is mostly complete with negligible pieces of data missing. I was able to filter those data out without dramatically altering the outcome of my graphs.

Taking a quick look through the top 20 receivers of grants for the last 9 years, it is interesting to note that the top 20 beneficiaries of Toronto City community grants are frequently the same few organizations. Organizations such as Toronto Symphony Orchestra, Toronto International Film Festival Inc(TIFF), Toronto Caribbean Carnival (Carabana) and Toronto Foundation for Student Success are almost among the top receivers among over 800 organizations to receive these grants every year. Furthermore, every year with the exception of 2019, Toronto Foundation for Student Success has lead the received amount by a significant amount. In 2019, the leader was Angel Foundation for Learning. Both these organizations received money from the Student Nutrition Program. 
```{r,figures-side, fig.show="hold", echo = FALSE}
graph_top_receivers(top_20_receivers_2019,"Top 20 Funding Receivers of 2019")
graph_top_receivers(top_20_receivers_2018,"Top 20 Funding Receivers of 2018")
graph_top_receivers(top_20_receivers_2017,"Top 20 Funding Receivers of 2017")
graph_top_receivers(top_20_receivers_2016,"Top 20 Funding Receivers of 2016")
graph_top_receivers(top_20_receivers_2015,"Top 20 Funding Receivers of 2015")
graph_top_receivers(top_20_receivers_2014,"Top 20 Funding Receivers of 2014")
graph_top_receivers(top_20_receivers_2013,"Top 20 Funding Receivers of 2013")
graph_top_receivers(top_20_receivers_2012,"Top 20 Funding Receivers of 2012")
graph_top_receivers(top_20_receivers_2011,"Top 20 Funding Receivers of 2011")
graph_top_receivers(top_20_receivers_2010,"Top 20 Funding Receivers of 2010")
```

Next I looked through the total funding that was approved by the city of Toronto every year, along with the total money received by the top 20 receivers. As expected, the general trend of the total funding is mimicked by the funding given to the top 20 organizations. However, there is a much steeper increase in the funding towards the top 20 organizations compared to the total funding given out ever year. The total funding given rose over 50 percent between 2010 to 2019, from roughly 43 million dollars to almost 69 million dollars. Despite the impressive increase, the amount of money the top 20 organizations received rose from 15 million dollars to almost 30 million dollars, almost a 100 percent increase in funding over 9 years. 

```{r, echo = FALSE}
ggplot(total_funding_by_year, aes(year, funding, group=type)) +
  geom_line(aes(colour = type), position = "stack") +
  scale_x_continuous("Year",breaks= pretty_breaks()) +
  scale_y_continuous("Funding Allocation ($)", labels = comma)

```
To investigate further, I looked into the percent break down of the funding each year. I analyzed how much percent the top 20 organizations receive in comparison the total amount of funds granted every year. Similar to the conclusions drawn in the graph previously.  In 2010, the top 20 companies received 33 percent of all the funds, and while a lot, in 2019 those funds rose to 43 percent of all the funds. 

```{r, echo = FALSE}

ggplot(total_funding_by_year_stacked,
       aes(x = year, y = funding, fill = type, label=round(funding))) + 
  geom_bar(stat="identity") + 
  geom_text(size = 3, position = position_stack(vjust = 0.5)) +
  scale_y_continuous("Funding Allocation (%)", labels = comma) + 
  scale_x_continuous(breaks= pretty_breaks())
```
Lastly, I quickly graphed all the amount of organizations that received the funding. Throughout the past 10 years, there were around 800 organizations that received funding every year. 

```{r, echo = FALSE}
ggplot(data = total_organization_by_year, aes(x = year, y = organizations )) +geom_bar(stat="identity")+ 
  scale_x_continuous(breaks= pretty_breaks())

```

# Discusion

The graphs showed a trend that as city increased funding, a significant portion of the increased funding went towards programs and organizations that already receive a significant portion of the grants. Almost all of the organizations in the top 20 list are well known and generally appreciated throughout the city, including a wide variety of arts programs, cultural festivities and youth support, supporting the cities push for diversity and inclusiveness. 

The city shows favor towards a few organizations by allocating a significant portion of the funds to these organizations every year. As the city increases the budget for the annual community grants, the increases goes towards more funding towards the well funded programs instead of newer programs, as the city has not significantly increased the number of organizations they fund every year. Despite well over 50 million dollars of funding being rejected in 2015 (City of Toronto, 2016), the city of Toronto has not made an attempt to provide the these organizations with grants. 

Summarizing the various graphs, while the city provides a lot of funds to that certainly help out many communities, the city provides by far the most money through the Student Nutrition Program through 2 organizations; the Angel Foundation for Learning (the main beneficiary in 2019) and the Toronto Foundation for Student Success (main beneficiary through 2010-2018). The heavy emphasis on these funding through the past 10 years show that these believes are widely supported by the city, and are irrelevant to the reigning members of city council. 

Full code and data can be found in the R Markdown file located [here](https://github.com/JamesBond0014/STA304_Blogdown/blob/master/content/post/2015-07-23-r-rmarkdown.Rmd).

# Weaknesses and next steps

The data sets between each year is slightly different, making comparing data between each year difficult and time consuming. Due to the issue, some more in depth analysis could not be done, such as grant distribution by ward or by funding programs. Next steps would be modifying the data sets in order to create meaningful graphs and analyze interesting trends regarding wards. 

# References

City of Toronto (2016). Impact of CPIP Grants. https://www.toronto.ca/legdocs/mmis/2017/cd/bgrd/backgroundfile-106491.pdf
  
Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2020).
dplyr: A Grammar of Data Manipulation. https://dplyr.tidyverse.org,
https://github.com/tidyverse/dplyr.

Hadley Wickham and Dana Seidel (2020). scales: Scale Functions for
Visualization. https://scales.r-lib.org,
https://github.com/r-lib/scales.

H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag
New York, 2016.

Sharla Gelfand (2020). opendatatoronto: Access the City of Toronto Open
Data Portal. https://sharlagelfand.github.io/opendatatoronto/,
https://github.com/sharlagelfand/opendatatoronto/.

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source
Software, 4(43), 1686, https://doi.org/10.21105/joss.01686

